---
title: "3. Preliminary Analysis"
author: "Jaejin Kim"
date: "4/1/2019"
output: pdf_document
---

```{r}
suppressPackageStartupMessages(require(dplyr))
suppressPackageStartupMessages(require(psych))
suppressPackageStartupMessages(require(devtools))
suppressPackageStartupMessages(require(ggplot2))
suppressPackageStartupMessages(require(openair))
suppressPackageStartupMessages(require(data.table))
suppressPackageStartupMessages(require(knitr))
```

# Load dataset and format date
```{r}
billboard <- read.csv("../Data/Clean/Billboard_spotify.csv")
billboard$time <- as.Date(billboard$time)
```

# Extract data to visualize the trend in music features
```{r}
#total danceability
total.dance<-aggregate(list(danceability = billboard$danceability),
                           list(time = cut(billboard$time, "1 year")),
                           mean)
total.dance$time <- as.Date(total.dance$time)

#total valence
total.val<-aggregate(list(valence = billboard$valence),
                           list(time = cut(billboard$time, "1 year")),
                           mean)
total.val$time <- as.Date(total.val$time)

#total energy
total.energy<-aggregate(list(energy = billboard$energy),
                           list(time = cut(billboard$time, "1 year")),
                           mean)
total.energy$time <- as.Date(total.energy$time)

#total loudness
total.loud<-aggregate(list(loudness = billboard$loudness),
                           list(time = cut(billboard$time, "1 year")),
                           mean)
total.loud$time <- as.Date(total.loud$time)

#total tempo
total.tempo<-aggregate(list(tempo = billboard$tempo),
                           list(time = cut(billboard$time, "1 year")),
                           mean)
total.tempo$time <- as.Date(total.tempo$time)
```

# Distribution of music features
```{r}
hist(total.val$valence, main = "Histogram of Valence", xlab = "Valence")
hist(total.dance$danceability, main = "Histogram of Danceability", xlab = "Danceability")
hist(total.loud$loudness, main = "Histogram of Loudness", xlab = "Loudness")
hist(total.tempo$tempo, main = "Histogram of tempo", xlab = "Tempo")
hist(total.energy$energy, main = "Histogram of Energy", xlab = "Energy")
```

```{r}
#plot danceability
ggplot(data=total.dance, aes(x=time, y=danceability)) +
  geom_line() +
  geom_hline(yintercept = mean(total.dance$danceability), linetype = "dashed", color ="red") +
  annotate(geom="text", label = "mean",
           x = as.Date("2020-01-01"), y = mean(total.dance$danceability),
           vjust = -1, color = "red") +
  geom_point() +
  ggtitle("Danceability, 1958 ~ 2018") +
  scale_x_date(date_labels = '%Y/%m')
```

```{r}
#plot valence
ggplot(data=total.val, aes(x=time, y=valence)) +
  geom_line() +
  geom_hline(yintercept = mean(total.val$valence), linetype = "dashed", color ="red") +
  annotate(geom="text", label = "mean",
           x = as.Date("2020-01-01"), y = mean(total.val$valence),
           vjust = -1, color = "red") +
  geom_point() +
  ggtitle("Valence, 1958 ~ 2018") +
  scale_x_date(date_labels = '%Y/%m')
```

```{r}
#plot energy
ggplot(data=total.energy, aes(x=time, y=energy)) +
  geom_line() +
  geom_hline(yintercept = mean(total.energy$energy), linetype = "dashed", color ="red") +
  annotate(geom="text", label = "mean",
           x = as.Date("2020-01-01"), y = mean(total.energy$energy),
           vjust = -1, color = "red") +
  geom_point() +
  ggtitle("Energy, 1958 ~ 2018") +
  scale_x_date(date_labels = '%Y/%m')
```

```{r}
#plot loudness
ggplot(data=total.loud, aes(x=time, y=loudness)) +
  geom_line() +
  geom_hline(yintercept = mean(total.loud$loudness), linetype = "dashed", color ="red") +
  annotate(geom = "text", label = "mean",
           x = as.Date("2020-01-01"), y = mean(total.loud$loudness),
           vjust = -1, color = "red") +
  geom_point() +
  ggtitle("Loudness, 1958 ~ 2018") +
  scale_x_date(date_labels = '%Y/%m')
```

```{r}
#plot tempo
ggplot(data=total.tempo, aes(x=time, y=tempo)) +
  geom_line() +
  geom_hline(yintercept = mean(total.tempo$tempo), linetype = "dashed", color ="red") +
  annotate(geom="text", label = "mean",
           x = as.Date("2020-01-01"), y = mean(total.tempo$tempo),
           vjust = -1, color = "red") +
  geom_point() +
  ggtitle("Tempo, 1958 ~ 2018") +
  scale_x_date(date_labels = '%Y/%m')
```


# Extract data from 2000 ~ 2014 to observe if there are any significant changes in music features during the financial crisis
```{r}
Fin.crisis <- subset(billboard,
                     format.Date(time, "%Y")=="2000" |
                     format.Date(time, "%Y")=="2001" |
                     format.Date(time, "%Y")=="2002" |
                     format.Date(time, "%Y")=="2003" |
                     format.Date(time, "%Y")=="2004" |
                     format.Date(time, "%Y")=="2005" |
                     format.Date(time, "%Y")=="2006" |
                     format.Date(time, "%Y")=="2007" |
                     format.Date(time, "%Y")=="2008" | 
                     format.Date(time, "%Y")=="2009" |
                     format.Date(time, "%Y")=="2010" |
                     format.Date(time, "%Y")=="2011" |
                     format.Date(time, "%Y")=="2012" |
                     format.Date(time, "%Y")=="2013" |
                     format.Date(time, "%Y")=="2014")

Fin.crisis$time <- as.Date(Fin.crisis$time)
```

# Danceability
```{r}
#create yearly average danceability
fin.dance.av <- aggregate(list(danceability = Fin.crisis$danceability),
                           list(time = cut(Fin.crisis$time, "1 year")),
                           mean)
fin.dance.av$time <- as.Date(fin.dance.av$time)

#plot
ggplot(data=fin.dance.av, aes(x=time, y=danceability)) +
  geom_line() +
  geom_point() +
  ggtitle("Danceability, 2000 ~ 2014") +
  scale_x_date(date_labels = '%Y/%m')
```

# Valence
```{r}
#create yearly average valence
fin.valence.av <- aggregate(list(valence = Fin.crisis$valence),
                           list(time = cut(Fin.crisis$time, "1 year")),
                           mean)
fin.valence.av$time <- as.Date(fin.valence.av$time)

#plot
ggplot(data=fin.valence.av, aes(x=time, y=valence)) +
  geom_line() +
  geom_point() +
  ggtitle("Valence, 2000 ~ 2014") +
  scale_x_date(date_labels = '%Y/%m')
```

# Energy
```{r}
#create yearly average energy
fin.energy.av <- aggregate(list(energy = Fin.crisis$energy),
                           list(time = cut(Fin.crisis$time, "1 year")),
                           mean)
fin.energy.av$time <- as.Date(fin.energy.av$time)

#plot
ggplot(data=fin.energy.av, aes(x=time, y=energy)) +
  geom_line() +
  geom_point() +
  ggtitle("Energy, 2000 ~ 2014") +
  scale_x_date(date_labels = '%Y/%m')
```

#loudness
```{r}
#create yearly average loudness
fin.loud.av <- aggregate(list(loud = Fin.crisis$loud),
                           list(time = cut(Fin.crisis$time, "1 year")),
                           mean)
fin.loud.av$time <- as.Date(fin.loud.av$time)

#plot
ggplot(data=fin.loud.av, aes(x=time, y=loud)) +
  geom_line() +
  geom_point() +
  ggtitle("Loudness, 2000 ~ 2014") +
  scale_x_date(date_labels = '%Y/%m')
```

#tempo
```{r}
#create yearly average tempo
fin.tempo.av <- aggregate(list(tempo = Fin.crisis$tempo),
                           list(time = cut(Fin.crisis$time, "1 year")),
                           mean)
fin.tempo.av$time <- as.Date(fin.tempo.av$time)

#month
ggplot(data=fin.tempo.av, aes(x=time, y=tempo)) +
  geom_line() +
  geom_point() +
  ggtitle("Tempo, 2000 ~ 2014") +
  scale_x_date(date_labels = '%Y/%m')
```



# What happened during the 1990s?
We see a huge drop in all of the music features except for loudness.
```{r}
nineties <- subset(billboard,
                   format.Date(WeekID, "%Y")=="1990" |
                   format.Date(WeekID, "%Y")=="1991" |
                   format.Date(WeekID, "%Y")=="1992" |
                   format.Date(WeekID, "%Y")=="1993" |
                   format.Date(WeekID, "%Y")=="1994" |
                   format.Date(WeekID, "%Y")=="1995" |
                   format.Date(WeekID, "%Y")=="1996" |
                   format.Date(WeekID, "%Y")=="1997" |
                   format.Date(WeekID, "%Y")=="1998" |
                   format.Date(WeekID, "%Y")=="1999")
```

```{r}
artist90s <- table(nineties$artist)
kable(head(sort(artist90s[artist90s!=0], decreasing = T), n = 100), format = "markdown")
```

```{r}
songs90s <- table(nineties$title)
kable(head(sort(songs90s[songs90s!=0], decreasing = T), n = 100), format = "markdown")
```

```{r}
hist(nineties$valence, main = "Histogram of Valence during the 90s", xlab = "Valence")
hist(nineties$danceability, main = "Histogram of Danceability during the 90s", xlab = "Danceability")
hist(nineties$energy, main = "Histogram of Energy during the 90s", xlab = "Valence")
hist(nineties$tempo, main = "Histogram of Tempo during the 90s", xlab = "Tempo")
hist(nineties$loudness, main = "Histogram of Loudness during the 90s", xlab = "Loudness")
```


# GDP growth rate
```{r}
GDP <- read.csv("../Data/Clean/GDP_cleaned.csv")
GDP$time <- as.Date(GDP$time)

ggplot(data=GDP, aes(x=time, y=gdp)) +
  geom_line() +
  geom_point() +
  ggtitle("GDP growth rate, 1961 ~ 2017") +
  scale_x_date(date_labels = '%Y') +
  annotate(geom="text", label = "mean",
           x = as.Date("2020-01-01"), y = mean(GDP$gdp),
           vjust = -1, color = "red") +
  geom_hline(yintercept = mean(GDP$gdp), linetype = "dashed", color ="red")
```







